## {background-color="#533860"}

::: {style="color: white; font-size: 1.8em; font-weight: bold;"}
💡 Random Thoughts
:::
::: {style="color: #FFE42A; font-size: 1.1em;"}
Things that don't fit elsewhere (but should)
:::

---

## Three Ways to Average {#three-means}

- **Arithmetic**: sum / n — the default
- **Geometric**: (x₁ × x₂ × ... × xₙ)^(1/n) → equivalently: `exp(mean(log(x)))`
- **Harmonic**: n / (1/x₁ + 1/x₂ + ... + 1/xₙ)

::: {.fragment}
**AM ≥ GM ≥ HM** — always true for positive numbers
:::

---

## When to Use Which {#which-mean}

:::: {.columns}

::: {.column width="50%"}
**Arithmetic**

- Avg height of students
- Avg test score
- Additive quantities
:::

::: {.column width="50%"}
**Geometric**

- Avg annual investment return
- Compounding quantities
- Growth rates

::: {.fragment}
+50% then −50%:

- Arithmetic avg = **0%**
- Geometric avg = **−13.4%** ← real answer
:::
:::

::::

::: {.fragment}
**Harmonic**

- Avg speed over equal *distances*
- Drive 60 mph there, 40 mph back → avg is **NOT** 50 mph
- 2 / (1/60 + 1/40) = **48 mph**
- Also: F1-score in ML (balances precision & recall)
:::

---

## A Surprising Example {#investment-example}

```{r}
#| echo: true
#| eval: false

# Invest $1000 over 2 years
# Year 1: +100% (doubles to $2000)
# Year 2: -50% (back to $1000)
arithmetic_avg <- mean(c(100, -50))             # = 25% — MISLEADING
geometric_avg  <- prod(c(2.0, 0.5))^(1/2) - 1  # = 0%  — correct!
```

::: {.fragment}
- Arithmetic mean says **25% average return**
- You ended up with **exactly what you started with**
- Geometric mean tells the truth
:::

---

## Should You Scale Dummy Variables? {#scale-dummies}

**Setup: one-hot encoding**

- Convert categorical → binary dummy variables
- "Color: Red/Blue/Green" → `is_red`, `is_blue`, `is_green`
- Each column is 0 or 1

::: {.fragment}
**The question**: after encoding, should you **center and scale** these binary columns?
:::

::: {.fragment}
Short answer: **it depends on the algorithm**
:::

---

## It Depends on the Algorithm {#scale-by-algo}

| Algorithm | Scale dummies? | Why |
|-----------|----------------|-----|
| **KNN** | ❌ No | Distances between 0/1 are already meaningful |
| **LASSO/Ridge** | ✓ Usually yes | Penalty should be scale-invariant |
| **Tree methods** | Doesn't matter | Split-based, scale-invariant |
| **Neural nets** | ✓ Yes | Gradient descent converges faster |

---

## The KNN Problem — Don't Scale {#knn-dummies}

2 features:

- `age` (scaled to 0–1 range)
- `is_rare_category` (mostly 0, rarely 1) — variance ≈ 0.02

::: {.fragment}
After standard scaling `(x - mean) / sd`:

- `age`: stays roughly 0–1 range
- `is_rare_category`: gets blown up to range ≈ **0–7!**
:::

::: {.fragment}
```{r}
#| echo: true
#| eval: false

set.seed(42)
n        <- 1000
age      <- runif(n, 18, 65)
rare_cat <- rbinom(n, 1, 0.05)   # variance ≈ 0.0475

age_scaled  <- scale(age)         # sd ≈ 1
rare_scaled <- scale(rare_cat)    # sd = 1, but range ≈ -0.22 to 4.5!
```

Scaling **inflates** the influence of low-variance dummies in KNN distance calculations
:::

---

## The LASSO Problem — Maybe Scale {#lasso-dummies}

LASSO penalty: λ × |β_j| for each coefficient

:::: {.columns}

::: {.column width="50%"}
**Without scaling**

- `income` in dollars → small βs
- `is_employed` in 0/1 → larger βs
- LASSO penalizes `is_employed` more
- Even if equally important
:::

::: {.column width="50%"}
::: {.fragment}
**With scaling**

- Penalty is "fair" across features
- Coefficients live on the same scale

**But:** low-variance dummies → scaling amplifies noise
:::
:::

::::

---

## The Low Variance Example {#low-variance-lasso}

```{r}
#| echo: true
#| eval: false

library(glmnet)
set.seed(1)
n <- 500

# True model: y = 2*x + 5*rare_event + noise
x    <- rnorm(n)
rare <- rbinom(n, 1, 0.02)   # only ~1% events
y    <- 2*x + 5*rare + rnorm(n)

X_unscaled <- cbind(x, rare)
X_scaled   <- scale(X_unscaled)

fit_u <- glmnet(X_unscaled, y, alpha = 1)
fit_s <- glmnet(X_scaled,   y, alpha = 1)
```

::: {.fragment}
- With only ~10 rare events, even scaling can't save you
- LASSO may shrink the rare coefficient to **0** — signal just isn't there
- Need more data, or a different approach
:::

---

[← Back to Topics](#toc)
